The imprint of global magma oceans on exoplanet demographics

### 2022-12-20 src/tex/ms.tex
introduce moving average, likelihood function

% Main body
\section{Introduction}
\todo[inline]{papers: \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021,Downey2022}.}

@@

\begin{note}
    As a null hypothesis, we consider the case where the planetary radius distribution is independent of the instellation,
    \begin{equation}
        h_{\mathrm{null}}(\theta, S) = \theta,
    \end{equation}
    where $\theta$ is the set of parameters defining the radius distribution.
    We further define an alternative hypothesis that describes radius changes due to steam atmospheres and water in the melt.
    As motivated above, this hypothesis takes the form of a step function in instellation $S$, where the step occurs at the outer edge of the runaway greenhouse region.
    Our main observable shall be the average planet radius in the planet population inside and outside this threshold.
    The magma ocean hypothesis is then defined as
\begin{equation}\label{eq:mo_hypo}
    h_{\mathrm{MO}}(\theta, S) =
%        \begin{cases}
%        1 & \text{if } x \in \mathbb{Q}\\
%        0 & \text{if } x \in \mathbb{R}\setminus\mathbb{Q}
%    \end{cases}
        \begin{cases}
        h_{\mathrm{null}} & \text{for } S \leq S_\mathrm{thresh}\\
        <R_\mathrm{P}> (f_\mathrm{steam}, f_\mathrm{water}) & \text{for } S > S_\mathrm{thresh}.
    \end{cases}
\end{equation}
    Here, $f_\mathrm{steam}$ and $f_\mathrm{water}$ are predictions from the steam atmosphere and water incorporation models, respectively.
    They are assumed to act additively on the planet radii and thus on their average $<R_\mathrm{P}>$.

%        \begin{cases}
%        \end{cases}

\begin{figure}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/HnullHmo.pdf}
        \caption{Planets above and below the runaway greenhouse threshold.
        \textit{Top}: Planet state as a function of orbital period. Planets with and without a runaway greenhouse climate mix and are not distinguishable in period space.
        \textit{Bottom}: Radii of synthetic planets with injected radius deviation as a function of instellation. Only the planets marked as transiting are observable. Above the runaway greenhouse threshold $S_\mathrm{thresh} =\SI{280}{\watt\per\meter\squared} $, radii are inflated by the amount indicated with gray lines. The sharp boundary at $S_\mathrm{thresh}$ causes a discontinuity in the average planet radius (blue). This magma ocean hypothesis can be tested agains the null hypothesis $H_0$ (green), where average radii are independent of instellation.}
        \label{fig:HnullHmo}
    \end{centering}
\end{figure}
    Figure~\ref{fig:HnullHmo} shows the null and alternative hypotheses as a function of instellation.
    The value for the average radius \textit{outside} the runaway greenhouse regime was computed from our nominal observed sample, whereas the radius change in the magma ocean case compared to the null hypothesis is a prediction from the magma ocean model, evaluated on the same sample.

...

For the likelihood function, we assumed that the planet radii $R_{\mathrm{P}, i}$ are measured with a normally distributed uncertainty $\sigma_{R_\mathrm{P}, i}$ and adopted a normal distribution
\begin{equation}
    \mathcal{L}(R_\mathrm{P} \mid \boldsymbol{\theta})=\prod_{i}^{N} \frac{1}{\sqrt{2 \pi \sigma_{R_\mathrm{P}, i}^{2}}} \exp \left(-\frac{\left(R_{\mathrm{P}, i}-h\left(\boldsymbol{\theta}, S_i\right)\right)^{2}}{2 \sigma_{R_\mathrm{P}, i}^{2}}\right).
\end{equation}
Here, $h\left(\boldsymbol{\theta}, S_i\right)$ corresponds to the functional form of the magma ocean hypothesis~(Eqn.~\ref{eq:mo_hypo}).
We tested this hypothesis against the null hypothesis $h_\mathrm{null} (\theta, R_\mathrm{P}) = \theta$, which states that there is no relationship between the measured planet radii and received instellation.

>>>>

% Main body
\section{Introduction}
\todo[inline]{papers: \citep[][introduce runaway GH]{} \citep[][statistical detection of GH transition]{Turbet2019}, \citep[][water loss, O2 buildup around M dwarfs]{Luger2015}, \citep[][Solubility of H in magma oceans]{Hirschmann2012,SchlichtingShahar_inreview,} \citet{Hamano2013,Hamano2015,Barth2021,Downey2022}.}

@@

\begin{note}
    As a null hypothesis, we consider the case where the planetary radius distribution is independent of the instellation,
    \begin{equation}
        h_{\mathrm{null}}(\theta, S) = \theta,
    \end{equation}
    where $\theta$ is the set of parameters defining the radius distribution.
    We further define an alternative hypothesis that describes radius changes due to steam atmospheres and water in the melt.
    As motivated above, this hypothesis takes the form of a step function in instellation $S$, where the step occurs at the outer edge of the runaway greenhouse region.
    Our main observable shall be the average planet radius in the planet population inside and outside this threshold.
    The magma ocean hypothesis is then defined as
\begin{equation}\label{eq:mo_hypo}
    h_{\mathrm{MO}}(\theta, S) =
%        \begin{cases}
%        1 & \text{if } x \in \mathbb{Q}\\
%        0 & \text{if } x \in \mathbb{R}\setminus\mathbb{Q}
%    \end{cases}
        \begin{cases}
        h_{\mathrm{null}} & \text{for } S \leq S_\mathrm{thresh}\\
        <R_\mathrm{P}> (f_\mathrm{steam}, f_\mathrm{water}) & \text{for } S > S_\mathrm{thresh}.
    \end{cases}
\end{equation}
    Here, $f_\mathrm{steam}$ and $f_\mathrm{water}$ are predictions from the steam atmosphere and water incorporation models, respectively.
    They are assumed to act additively on the planet radii and thus on their average $<R_\mathrm{P}>$.
\begin{figure}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/HnullHmo.pdf}
        \caption{Planets above and below the runaway greenhouse threshold.
        \textit{Top}: Planet state as a function of orbital period. Planets with and without a runaway greenhouse climate mix and are not distinguishable in period space.
        \textit{Bottom}: Radii of synthetic planets with injected radius deviation as a function of instellation. Only the planets marked as transiting are observable. Above the runaway greenhouse threshold $S_\mathrm{thresh} =\SI{280}{\watt\per\meter\squared} $, radii are inflated by the amount indicated with gray lines. The sharp boundary at $S_\mathrm{thresh}$ causes a discontinuity in the average planet radius (blue). This magma ocean hypothesis can be tested agains the null hypothesis $H_0$ (green), where average radii are independent of instellation.}
        \label{fig:HnullHmo}
    \end{centering}
\end{figure}
    Figure~\ref{fig:HnullHmo} shows the null and alternative hypotheses as a function of instellation.
    The value for the average radius \textit{outside} the runaway greenhouse regime was computed from our nominal observed sample, whereas the radius change in the magma ocean case compared to the null hypothesis is a prediction from the magma ocean model, evaluated on the same sample.
\end{note}

\begin{note}
The measured radii $R_\mathrm{P, i}$ cannot be directly used for the hypothesis tests as they include intrinsic scatter that is not caused by measurement errors.
$h_{\mathrm{MO}}$ and $h_{\mathrm{null}}$ should thus be tested against a statistical estimator that represents the population mean.
To avoid binning and the artificial patterns it may introduce, we chose to test our hypotheses against a simple moving average $SMA$ along the instellation axis with a window of size \windowsize\ centered around each measurement.
We further computed the uncertainty of this moving average by propagating the individual measurement errors and applying a rolling standard error of the mean. \todo[inline]{specify error propagation}

As our procedure involves random sampling of the model parameters $\theta$, we need to define the probability of obtaining a dataset given the model parameters, i.e., a likelihood function.
We assumed here that the individual moving averages $SMA_i$ are measured with a normally distributed uncertainty $\sigma_{SMA_i}$ and adopted a normal distribution
\begin{equation}
    \mathcal{L}(SMA \mid \boldsymbol{\theta})=\prod_{i}^{N} \frac{1}{\sqrt{2 \pi \sigma_{SMA_i}^{2}}} \exp \left(-\frac{\left(SMA_i - h\left(\boldsymbol{\theta}, S_i\right)\right)^{2}}{2 \sigma_{SMA_i}^{2}}\right).
\end{equation}
Here, $h\left(\boldsymbol{\theta}, S_i\right)$ corresponds to the functional form of the magma ocean hypothesis~(Eqn.~\ref{eq:mo_hypo}).
We tested this hypothesis against the null hypothesis $h_\mathrm{null} (\theta, R_\mathrm{P}) = \theta$, which states that there is no relationship between the measured planet radii and received instellation.
